Grappa: Scalable Computing on Mass-Market Hardware for Irregular, Low-Locality Applications

### 2012-07-22 doc/asplos13/relatedwork.tex
work around bad citation

\section{Related work}

Ways to organize this:
Approaches to graphs: special purpose XMT -> cluster -> distributed frameworks {pregel,graphlab} -> limited programming model -> grappa gen purpose

dsm, pgas
lightweight threading
other graph solutions: 
-graph parallel, distributed
--pregel - but is synchronous and thus inefficient
--graphlab - asynchrounous, esp useful for problems that have been shown empirically to converge faster with asynchrony

\subsection{Multithreading to tolerate latency}
hardware: tera, cray
smt,
alewife
gpu,
software threads

\subsection{Programming models for distributed graph processing}

Distributed graph processing frameworks like Pregel \cite{pregel:2010} and Distributed GraphLab \cite{distgraphlab:vldb12} provide graph-parallel, vertex-centric programming abstractions that free the application writer from solving distributed system issues like scheduling parallelism, handling fault tolerance, and scaling communication. Pregel adopts a bulk-synchronous parallel (BSP) execution model, which makes it inefficient on workloads that could prioritize vertices. GraphLab, on the other hand, schedules vertex computations individually, allowing prioritization, which gives faster convergence in a variety of iterative algorithms. Grappa also supports dynamic parallelism with asynchronous execution, but has a more general purpose programming model, where parallelism is expressed as tasks or loop iterations. \todo{and express locality? mention pgas e.g. Chapel language implementation? How much to talk about how graphlab achieves scalable performance?}

\subsection{Partitioned Global Address space}
The goal of presenting a global view of distributed memory to the programmer is shared by the PGAS community, and is used in languages like Chapel \cite{Chapel}, X10 \cite{X10}, and UPC \cite{UPC}. In these programming models, access to shared structures look like normal memory references, but the programmer tries to minimize references to remote nodes. Unlike the usual programming goal in PGAS, since we target problems with poor locality, we design Grappa for remote references as a common case. Thread private data is still local, and unlike the XMT, we exploit locality where it does exist in an application (such as spatial locality in a graph edgelist), but we design the system to perform well in the midst of mostly remote accesses to large shared data structures. We would like to implement PGAS languages that support dynamic parallelism, like Chapel, in Grappa.

>>>>

\section{Background}

The goal of the Grappa framework is to simplify the task of implementing irregular applications.  Grappa is built on many existing ideas in programming languages, systems and architecture.  In this section we discuss related frameworks and key enabling technologies that Grappa builds upon.

\subsection{Comparative Frameworks}

Distributed graph processing frameworks like Pregel \cite{pregel:2010} and Distributed GraphLab \cite{distgraphlab:vldb12} share similar goals.  Pregel adopts a bulk-synchronous parallel (BSP) execution model, which makes it inefficient on workloads that could prioritize vertices. GraphLab, on the other hand, schedules vertex computations individually, allowing prioritization, which gives faster convergence in a variety of iterative algorithms. Grappa also supports dynamic parallelism with asynchronous execution, but parallelism is expressed as tasks or loop iterations.

\subsection{Global memory}

Grappa implements a software distributed shared memory DSM system.  Many traditional software DSM systems are page based~\cite{Treadmarks} and aim to hide the fact that they are built in software from applications by exploiting the processor's paging mechanisms.  Instead Grappa, like other partitioned global address space (PGAS) models, implements its DSM at the language, rather than system level.  Languages such as Chapel \cite{Chapel}, X10 \cite{X10}, and UPC \cite{UPC} make accesses to shared structures look like normal memory references.  As we describe later, Grappa chooses a middle ground, where global addresses are explicit in the API and local accesses are emitted conventionally by the compiler.  Similar to DSM systems Grappa provides a caching and coherence model to ensure high performance and consistent results.  Cache operations are exposed to in the API, however, allowing software to optimize their usage.

\subsection{Multithreading}

Grappa uses multithreading to tolerate memory latency.  This is a well known technique.  Hardware implementations include the Tera MTA \cite{Tera}, Cray XMT \cite{XMT}, Simultaneous multithreading \cite{SMT}, MIT Alewife \cite{Alewife}, Cyclops \cite{Cyclops}, and even GPUs \cite{fatahalian}.  As we describe in this paper, Grappa use a lightweight user-mode task scheduler to multiplex thousands of tasks on a single processing core.  The large number of tasks is required because of the extremely large internode latency.
